---
title: "Line shopping and devigging in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Line shopping and devigging in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# The whole vignette runs against ParlayAPI's keyless sandbox
# endpoints. If this machine is offline (or the API is unreachable),
# skip evaluation so the package still builds cleanly.
can_run <- requireNamespace("curl", quietly = TRUE) &&
  curl::has_internet() &&
  tryCatch(
    nrow(parlayapiR::pa_sports(sandbox = TRUE)) > 0,
    error = function(e) FALSE
  )
knitr::opts_chunk$set(eval = can_run)
```

Two habits separate profitable bettors from the rest: shopping every
line across books before betting, and removing the bookmaker margin
(the "vig") before comparing a price to a model. This vignette does
both with `parlayapiR`, entirely against the keyless sandbox
endpoints, so you can run every chunk without an API key.

```{r setup}
library(parlayapiR)
```

## 1. Pull a board of odds

Sandbox mode serves synthetic but structurally identical data from
`/v1/sandbox/` endpoints. Prices come back in American format.

```{r}
events <- pa_odds("basketball_nba", markets = "h2h", sandbox = TRUE)
events[, c("home_team", "away_team", "commence_time")]
```

Each event carries a `bookmakers` list-column in the same shape used
by The Odds API v4: bookmakers, then markets, then outcomes.

## 2. Flatten to one row per quote

```{r}
quotes <- do.call(rbind, lapply(seq_len(nrow(events)), function(i) {
  ev <- events[i, ]
  books <- ev$bookmakers[[1]]
  do.call(rbind, lapply(seq_len(nrow(books)), function(b) {
    mkt <- books$markets[[b]][1, ]
    outs <- mkt$outcomes[[1]]
    data.frame(
      event = paste(ev$away_team, "at", ev$home_team),
      book = books$key[b],
      outcome = outs$name,
      american = outs$price
    )
  }))
}))
quotes$decimal <- pa_american_to_decimal(quotes$american)
head(quotes)
```

## 3. Line shopping: the best price per outcome

The whole point of aggregating 30+ sportsbooks is that someone is
almost always off consensus. Take the best available decimal price for
every outcome:

```{r}
best <- aggregate(decimal ~ event + outcome, data = quotes, FUN = max)
names(best)[3] <- "best_decimal"
best <- merge(
  best,
  quotes,
  by.x = c("event", "outcome", "best_decimal"),
  by.y = c("event", "outcome", "decimal")
)
best[, c("event", "outcome", "book", "best_decimal")]
```

## 4. Devig a reference book to get fair probabilities

A quoted -110 / -110 market implies 52.4% for each side, which sums
to 104.8%: the extra 4.8% is the margin. `pa_devig()` removes it.
Here we devig one book's two-way market three ways:

```{r}
ev1 <- quotes[quotes$event == quotes$event[1], ]
ref <- ev1[ev1$book == ev1$book[1], ]
ref

pa_devig(ref$american, odds_format = "american")
pa_devig(ref$american, odds_format = "american", method = "additive")
pa_devig(ref$american, odds_format = "american", method = "power")
```

The multiplicative method scales everything by the same factor. The
power method shrinks long shots harder, which better matches the
favorite-longshot bias in real markets. For two-way markets near even
money the three methods agree closely.

## 5. Edge: fair probability versus the best available price

An edge exists when the best shopped price implies a lower probability
than your fair estimate:

```{r}
fair <- pa_devig(ref$american, odds_format = "american", method = "power")
comp <- data.frame(
  outcome = ref$outcome,
  fair_prob = fair$fair_prob,
  best_decimal = best$best_decimal[match(
    paste(ev1$event[1], ref$outcome),
    paste(best$event, best$outcome)
  )]
)
comp$edge <- comp$fair_prob * comp$best_decimal - 1
comp
```

A positive `edge` means the shopped price beats the devigged fair
price. On real (non-sandbox) data this is exactly how you find value
created by one book lagging the market.

## 6. Size the bet with the Kelly criterion

```{r}
plays <- comp[comp$edge > 0, ]
if (nrow(plays)) {
  plays$stake <- pa_kelly(
    prob = plays$fair_prob,
    odds = plays$best_decimal,
    fraction = 0.25,
    bankroll = 1000
  )
}
plays
```

Quarter Kelly (`fraction = 0.25`) is the standard practical choice:
full Kelly assumes your probabilities are exactly right, and they
never are.

## Going live

Everything above works identically against live data. Get a free key
(1,000 credits per month, no card required) at
[parlay-api.com](https://parlay-api.com), then:

```{r, eval = FALSE}
pa_auth("your-api-key")
events <- pa_odds("basketball_nba", markets = "h2h,spreads,totals")
props <- pa_props("basketball_nba", markets = "player_points")
closes <- pa_closing_odds("basketball_nba", season = "2024")
```

Historical odds reach back to 2005, and closing lines include more
than 30 million player prop closes since 2022, which makes
`pa_closing_odds()` the natural calibration target for any model you
backtest in R.
